# Project: BRFSS Diabetes Risk Analysis
# Purpose: End to end analysis of  public health survey data,
# including data cleaning, feature selection, modeling, and reporting.
# Tools: R, tidyverse, caret, glmnet, ggplot2

---
title: "Diabetes probability based on behavioral, medical, demographic, and economic factors
       MGT 6203: Data Analytics in Business"
author: Hande Pehlivan
date: "2023-04-11"
output: html_document
---

```{r setup, include=FALSE}
#read the file
R.Version() # using R version 4.2.2
library(foreign)
library(DataExplorer)
library(DataExplorer)
library(glmnet)
library(tidyverse)
library(Hmisc)
library(pastecs)
library(ggplot2)
library(car)
library(class)
library(grid)
library(dplyr)
library(tidyr)
library(tidyverse)
library(gridExtra)
library(corrplot)
library(caret)
library(performance)
library("readxl")
library(RColorBrewer)
library(dplyr)
library(ROCR)
library(e1071) 
library(rpart)
library(kknn)
library(imbalance)
library(caTools)

```

```{r cars}
#please put LLCP2021.XPT file in the same directory as the currently running R script
df <- read.xport("LLCP2021.XPT")
#check column and row
print(paste('number of column is',ncol(df)))
print(paste('number of row',nrow(df)))



#receive the data where 85% of not missing
percent_non_missing <- function(t)
{
  mean(x = is.na(x = t)) < 0.15
}

df2<-Filter(f = percent_non_missing,
            x = df)



#check column and row
print(paste('number of column is',ncol(df2)))
print(paste('number of row',nrow(df2)))
```

```{r }
#just get the row related to diabetes
df3<- df2[which(df2$DIABETE4 %in% c('1','2','3','4')),]
head(df3)

#create target variables
df3$DIABETES <- with(df3, ifelse(DIABETE4 =='2'| DIABETE4 =='3',0, 1))
#check distinct value for DIABETE4
table(df3$DIABETES)


# there are same variables and they don't have importance,removing them and DIABETE4 since we have created new column
df_new<-subset(df3, select = -c(IDATE,IMONTH,IDAY,IYEAR,SEQNO,FMONTH,DISPCODE,X_PSU,X_MISFRT1,X_MISVEG1,X_FRUITE1,
                                X_FRTRES1,X_VEGRES1,X_VEGETE1,DIABETE4))

head(df_new)


#visualize the table
# as we see plot is flagging 0.15 of data missing as bad
plot_missing(df_new,
             title='Diabetes Dataset Missing Data',
             group=list("Excellent" = 0.0, "Good" = .06, "Ok" = .1, "Bad" = .15),
             theme_config = list(legend.position = c("top")))


#remove all rows with NA
df_final<-df_new[complete.cases(df_new),]

#re-check column and row
print(paste('number of column is',ncol(df_final)))
print(paste('number of row',nrow(df_final)))

#re-visualize the table
plot_missing(df_final,
             title='Diabetes Dataset Missing Data',
             group=list("Excellent" = 0.0, "Good" = .06, "Ok" = .1, "Bad" = .15),
             theme_config = list(legend.position = c("top")))

# check unique data types in the data frame
unique(sapply(df_final, class))

#all of them are numeric.Since we don't have any categorical variable, we don't need to exclude anything while scaling the data.
#we want to see which variables are important based on lasso selection.

scaled_df = as.data.frame(scale(df_final[,-124]))
df_scaled <- cbind(scaled_df,df_final[,124]) 
colnames(df_scaled)

colnames(df_scaled)[124] <- "DIABETES"

#recheck column names
colnames(df_scaled)

XP=data.matrix(df_scaled[,-124])
YP=data.matrix(df_scaled$DIABETES)


lasso=cv.glmnet(XP,YP,alpha=1,
                nfolds = 5,type.measure="mse",family="gaussian")


#Output the coefficients of the variables selected by lasso
coef(lasso, s=lasso$lambda.min)


# Run PCA on matrix of scaled predictors
pca <- prcomp(df_final[,1:123], scale. = TRUE)
summary(pca)


#check screen
screeplot(pca, type="lines",col="blue")

var <- pca$sdev^2
propvar <- var/sum(var)

plot(propvar, xlab = "Principal Component", ylab = "Proportion of Variance Explained",ylim = c(0,1), type = "b")

plot(cumsum(propvar), xlab = "Principal Component", ylab = "Cumulative Proportion of Variance Explained",
     ylim = c(0,1), type = "b")

# Creating a data frame of response variable and PCs
PCDiabetes <- as.data.frame(cbind(pca$x, df_final[,124]))
colnames(PCDiabetes)[124] <- "DIABETES"



s<-cor(df_final)
s

#check index for diabetes
grep("DIABETES", colnames(df_final))

#convert data frame
result<- round(s[,124],2)


#based on 0.07  check positive and negative correlation
first<-Filter(function(x) any(x > 0.07), result)

p_correlation<-names(result)[sapply(result, function(x) min(x))>=0]
p_correlation


n_correlation<-names(result)[sapply(result, function(x) min(x))<0]
n_correlation

second<-Filter(function(x) any(x < -0.07 ), result[n_correlation])
second




```

```{r}
# combine the data with the variables that are correlated and use some of researches to identify important variables
#use codebook to check the explanation of each variables
# https://www.cdc.gov/pcd/issues/2019/19_0109.htm
#https://www.cdc.gov/brfss/annual_data/2021/pdf/codebook21_llcp-v2-508.pdf
#sex,CHCKDNY2 are added based on the research

final_data<-subset(df_final, select = c(GENHLTH,EXERANY2,EMPLOY1,CHILDREN,ALCDAY5,X_PHYS14D,X_HCVU652,X_TOTINDA,X_RFHYPE6,X_RFCHOL3,X_AGEG5YR,
                                        X_AGE65YR,X_AGE80,X_AGE_G,WTKG3,X_BMI5,X_BMI5CAT,X_RFBMI5,DRNKANY5,PHYSHLTH,CHECKUP1,BPHIGH6,CHOLCHK3,
                                        TOLDHI3,CHOLMED3,CVDINFR4,CVDCRHD4,HAVARTH5,EDUCA,DIFFWALK,PNEUVAC4,QSTVER,X_MICHD,X_DRDXAR3,
                                        X_LMTACT3,X_LMTWRK3,X_CHLDCNT,X_EDUCAG,X_INCOMG1,X_STATE,X_RACE,DIABETES,SEXVAR,CHCKDNY2))



#re-check column and row
print(paste('number of column is',ncol(final_data)))
print(paste('number of row',nrow(final_data)))

# identify the multicollinearity between the remaining variables
cr<-cor(final_data)
cr

#check col names
colnames(final_data)


# find the correlation > 0.70

for (i in 1:nrow(cr)){
  correlations <-  which((cr[i,] > 0.70) & (cr[i,] != 1))
  
  if(length(correlations)> 0){
    print(colnames(final_data)[i])
    print(correlations)
  }
}

# finally we picked the final variables based on the correlation and research
# also we decided to eliminated the records where no answer,don't know or refused
# individually checked all the variables distinct values


library(tidyverse)
final_data <- final_data %>%
  filter(GENHLTH %in% c(1,2,3,4,5)) %>%
  filter(X_TOTINDA %in% c(1,2)) %>% 
  filter(EMPLOY1 %in% c(1,2,3,4,5,6,7,8)) %>%
  filter(CHILDREN %in% c(1,2,3,4,5,6,7,8,9,10,11,12,14,15,16,17,20,23,82,83,88 )) %>%
  filter(DRNKANY5 %in% c(1,2)) %>%
  filter(X_HCVU652 %in% c(1,2)) %>%
  filter(CHCKDNY2 %in% c(1,2)) %>%
  filter(X_AGE_G %in% c(1,2,3,4,5,6)) %>%
  filter(X_BMI5CAT %in% c(1,2,3,4)) %>%
  filter(CHECKUP1 %in% c(1,2,3,4,8)) %>%
  filter(BPHIGH6 %in% c(1,2,3,4)) %>%
  filter(CHOLMED3 %in% c(1,2)) %>%
  filter(HAVARTH5 %in% c(1,2)) %>%
  filter(X_EDUCAG %in% c(1,2,3,4)) %>%
  filter(DIFFWALK %in% c(1,2)) %>%
  filter(PNEUVAC4 %in% c(1,2)) %>%
  filter(X_MICHD %in% c(1,2)) %>%
  filter(X_INCOMG1 %in% c(1,2,3,4,5,6,7)) %>%
  filter(X_RACE %in% c(1,2,3,4,5,6,7,8)) 


# transform children
final_data$CHILDREN <- with(final_data, ifelse(CHILDREN ==88,0, 1))
#check distinct value for CHILDREN
table(final_data$CHILDREN)


#create final table
diabetes_data<-subset(final_data, 
                      select = c(
                        X_STATE,
                        X_RACE,
                        SEXVAR,
                        GENHLTH,
                        X_TOTINDA,
                        EMPLOY1,
                        CHILDREN ,
                        DRNKANY5,
                        X_HCVU652,
                        CHCKDNY2,
                        X_AGE_G,
                        X_BMI5CAT,
                        CHECKUP1,
                        BPHIGH6,
                        CHOLMED3,
                        HAVARTH5,
                        X_EDUCAG,
                        DIFFWALK,
                        PNEUVAC4,
                        X_MICHD,
                        X_INCOMG1,
                        DIABETES))


#re-check column and row
print(paste('number of column is',ncol(diabetes_data)))
print(paste('number of row',nrow(diabetes_data)))




# rename columns
colnames(diabetes_data)

names(diabetes_data)[which(names(diabetes_data) == "X_STATE")] <- "STATE_code"
names(diabetes_data)[which(names(diabetes_data) == "X_RACE")] <- "Race"
names(diabetes_data)[which(names(diabetes_data) == "SEXVAR")] <- "Gender"
names(diabetes_data)[which(names(diabetes_data) == "GENHLTH")] <- "GeneralHealth"
names(diabetes_data)[which(names(diabetes_data) == "X_TOTINDA")] <- "PhysicalActivity"
names(diabetes_data)[which(names(diabetes_data) == "EMPLOY1")] <- "Employment"
names(diabetes_data)[which(names(diabetes_data) == "CHILDREN")] <- "Children"
names(diabetes_data)[which(names(diabetes_data) == "DRNKANY5")] <- "Alcohol"
names(diabetes_data)[which(names(diabetes_data) == "X_HCVU652")] <- "HealthInsurance"
names(diabetes_data)[which(names(diabetes_data) == "CHCKDNY2")] <- "Kidney"
names(diabetes_data)[which(names(diabetes_data) == "X_AGE_G")] <- "Age"
names(diabetes_data)[which(names(diabetes_data) == "X_BMI5CAT")] <- "BMI"
names(diabetes_data)[which(names(diabetes_data) == "CHECKUP1")] <- "RoutineCheckup"
names(diabetes_data)[which(names(diabetes_data) == "BPHIGH6")] <- "BloodPressure"
names(diabetes_data)[which(names(diabetes_data) == "CHOLMED3")] <- "Cholestero"
names(diabetes_data)[which(names(diabetes_data) == "HAVARTH5")] <- "Arthritis"
names(diabetes_data)[which(names(diabetes_data) == "X_EDUCAG")] <- "Education"
names(diabetes_data)[which(names(diabetes_data) == "DIFFWALK")] <- "DifficultyWalking"
names(diabetes_data)[which(names(diabetes_data) == "PNEUVAC4")] <- "Pneumonia"
names(diabetes_data)[which(names(diabetes_data) == "X_MICHD")] <- "MI"
names(diabetes_data)[which(names(diabetes_data) == "X_INCOMG1")] <- "Income"
names(diabetes_data)[which(names(diabetes_data) == "DIABETES")] <- "Diabetes"
```

```{r }
head(diabetes_data)
describe(diabetes_data)
stat.desc(diabetes_data)

table(diabetes_data$Diabetes)
tail(diabetes_data)

coul <- brewer.pal(5, "Set2") 
par(mfrow=c(2,2))
counts <- table(diabetes_data$Diabetes,diabetes_data$STATE_code)
barplot(counts, main="STATE_code vs Diabetes",xlab="STATE_code",ylab="Count", col=coul,
        legend = rownames(counts))

counts <- table(diabetes_data$Diabetes,diabetes_data$Race)
barplot(counts, main="Race vs Diabetes",xlab="Race",ylab="Count", col=coul,
        legend = rownames(counts))

counts <- table(diabetes_data$Diabetes,diabetes_data$Gender)
barplot(counts, main="Gender vs Diabetes",xlab="Gender",ylab="Count", col=coul,
        legend = rownames(counts))

counts <- table(diabetes_data$Diabetes,diabetes_data$GeneralHealth)
barplot(counts, main="GeneralHealth vs Diabetes",xlab="GeneralHealth",ylab="Count", col=coul,
        legend = rownames(counts))
par(mfrow=c(2,2))
counts <- table(diabetes_data$Diabetes,diabetes_data$PhysicalActivity)
barplot(counts, main="PhysicalActivity vs Diabetes",xlab="PhysicalActivity",ylab="Count", col=coul,
        legend = rownames(counts))

counts <- table(diabetes_data$Diabetes,diabetes_data$Employment)
barplot(counts, main="Employment vs Diabetes",xlab="Employment",ylab="Count", col=coul,
        legend = rownames(counts))

counts <- table(diabetes_data$Diabetes,diabetes_data$Children)
barplot(counts, main="Children vs Diabetes",xlab="Children",ylab="Count", col=coul,
        legend = rownames(counts))

counts <- table(diabetes_data$Diabetes,diabetes_data$Alcohol)
barplot(counts, main="Alcohol vs Diabetes",xlab="Alcohol",ylab="Count", col=coul,
        legend = rownames(counts))
par(mfrow=c(2,2))

counts <- table(diabetes_data$Diabetes,diabetes_data$HealthInsurance)
barplot(counts, main="HealthInsurance vs Diabetes",xlab="HealthInsurance",ylab="Count", col=coul,
        legend = rownames(counts))
counts <- table(diabetes_data$Diabetes,diabetes_data$Kidney)
barplot(counts, main="Kidney vs Diabetes",xlab="Kidney",ylab="Count", col=coul,
        legend = rownames(counts))

counts <- table(diabetes_data$Diabetes,diabetes_data$Age)
barplot(counts, main="Age vs Diabetes",xlab="Age",ylab="Count", col=coul,
        legend = rownames(counts))

counts <- table(diabetes_data$Diabetes,diabetes_data$BMI)
barplot(counts, main="BMI vs Diabetes",xlab="BMI",ylab="Count", col=coul,
        legend = rownames(counts))
par(mfrow=c(2,2))

counts <- table(diabetes_data$Diabetes,diabetes_data$RoutineCheckup)
barplot(counts, main="RoutineCheckup vs Diabetes",xlab="RoutineCheckup",ylab="Count", col=coul,
        legend = rownames(counts))

counts <- table(diabetes_data$Diabetes,diabetes_data$BloodPressure)
barplot(counts, main="BloodPressure vs Diabetes",xlab="BloodPressure",ylab="Count", col=coul,
        legend = rownames(counts))

counts <- table(diabetes_data$Diabetes,diabetes_data$Cholestero)
barplot(counts, main="Cholestero vs Diabetes",xlab="Cholestero",ylab="Count", col=coul,
        legend = rownames(counts))
counts <- table(diabetes_data$Diabetes,diabetes_data$Arthritis)
barplot(counts, main="Arthritis vs Diabetes",xlab="Arthritis",ylab="Count", col=coul,
        legend = rownames(counts))
par(mfrow=c(2,2))

counts <- table(diabetes_data$Diabetes,diabetes_data$Education)
barplot(counts, main="Education vs Diabetes",xlab="Education",ylab="Count", col=coul,
        legend = rownames(counts))
counts <- table(diabetes_data$Diabetes,diabetes_data$DifficultyWalking)
barplot(counts, main="DifficultyWalking vs Diabetes",xlab="DifficultyWalking",ylab="Count", col=coul,
        legend = rownames(counts))

counts <- table(diabetes_data$Diabetes,diabetes_data$Pneumonia)
barplot(counts, main="Pneumonia vs Diabetes",xlab="Pneumonia",ylab="Count", col=coul,
        legend = rownames(counts))
counts <- table(diabetes_data$Diabetes,diabetes_data$MI)
barplot(counts, main="MI vs Diabetes",xlab="MI",ylab="Count", col=coul,
        legend = rownames(counts))
par(mfrow=c(2,1))

counts <- table(diabetes_data$Diabetes,diabetes_data$Income)
barplot(counts, main="Income vs Diabetes",xlab="Income",ylab="Count", col=coul,
        legend = rownames(counts))

#correlation plot
corr <- cor(diabetes_data)
corrplot(corr,method="circle")

#box plot 
attach(diabetes_data)
par(mfrow=c(2,2))
boxplot(STATE_code~Diabetes, main="STATE_code vs Diabetes",xlab="STATE_code",ylab="Diabetes", col=coul
        )

boxplot(Race~Diabetes, main="Race vs Diabetes",xlab="Race",ylab="Diabetes", col=coul
        )

boxplot(Gender~Diabetes, main="Gender vs Diabetes",xlab="Gender",ylab="Diabetes", col=coul
       )

boxplot(GeneralHealth~Diabetes, main="GeneralHealth vs Diabetes",xlab="GeneralHealth",ylab="Diabetes", col=coul
        )
par(mfrow=c(2,2))
boxplot(PhysicalActivity~Diabetes, main="PhysicalActivity vs Diabetes",xlab="PhysicalActivity",ylab="Diabetes", col=coul
        )

boxplot(Employment~Diabetes, main="Employment vs Diabetes",xlab="Employment",ylab="Diabetes", col=coul
      )

boxplot(Children~Diabetes, main="Children vs Diabetes",xlab="Children",ylab="Diabetes", col=coul
        )
boxplot(Alcohol~Diabetes, main="Alcohol vs Diabetes",xlab="Alcohol",ylab="Diabetes", col=coul
        )
par(mfrow=c(2,2))

boxplot(HealthInsurance~Diabetes, main="HealthInsurance vs Diabetes",xlab="HealthInsurance",ylab="Diabetes", col=coul
        )

boxplot(Kidney~Diabetes, main="Kidney vs Diabetes",xlab="Kidney",ylab="Diabetes", col=coul
)
boxplot(Age~Diabetes, main="Age vs Diabetes",xlab="Age",ylab="Diabetes", col=coul
        )

boxplot(BMI~Diabetes, main="BMI vs Diabetes",xlab="BMI",ylab="Diabetes", col=coul
        )
par(mfrow=c(2,2))

boxplot(RoutineCheckup~Diabetes, main="RoutineCheckup vs Diabetes",xlab="RoutineCheckup",ylab="Diabetes", col=coul
        )

boxplot(BloodPressure~Diabetes, main="BloodPressure vs Diabetes",xlab="BloodPressure",ylab="Diabetes", col=coul
       )

boxplot(Cholestero~Diabetes, main="Cholestero vs Diabetes",xlab="Cholestero",ylab="Diabetes", col=coul)
boxplot(Arthritis~Diabetes, main="Arthritis vs Diabetes",xlab="Arthritis",ylab="Diabetes", col=coul
        )
par(mfrow=c(2,2))

boxplot(Education~Diabetes, main="Education vs Diabetes",xlab="Education",ylab="Diabetes", col=coul
        )
boxplot(DifficultyWalking~Diabetes, main="DifficultyWalking vs Diabetes",xlab="DifficultyWalking",ylab="Diabetes", col=coul
       )

boxplot(Pneumonia~Diabetes, main="Pneumonia vs Diabetes",xlab="Pneumonia",ylab="Diabetes", col=coul
       )
boxplot(MI~Diabetes, main="MI vs Diabetes",xlab="MI",ylab="Diabetes", col=coul
        )
par(mfrow=c(2,1))

boxplot(Income~Diabetes, main="Income vs Diabetes",xlab="Income",ylab="Diabetes", col=coul)


```

```{r }

# MERGE THE DATASETS
Cost_of_living <- read_excel("Cost_of_living.xlsx")
head(Cost_of_living)

diabetes_data$Diabetes <- as.numeric(diabetes_data$Diabetes)

#Group by state, calculates number of diabetes for each state. 

state_diabetes <- diabetes_data %>%
  group_by(STATE_code) %>%
  summarise(total_diabetes = sum(Diabetes))



# keep desired columns
print(state_diabetes)

merged_data <- merge(state_diabetes, Cost_of_living, by ="STATE_code")
head(merged_data)

#Population Data
Population <- read.csv("Population.csv", header=TRUE)
new_data <- Population[, c("state", "densityMi")]
names(new_data)[which(names(new_data) == "state")] <- "State"
merged_data2 <- merge(merged_data, new_data, by ="State")

#Create the diabetes Ratio = Diabetes/Density
merged_data2$Diabetes_Ratio <- (merged_data2$total_diabetes / merged_data2$densityMi)
head(merged_data2)

#Remove Alaska because it is outlier. 
merged_data2 <- merged_data2[merged_data2$State != "Alaska",]


ggplot(merged_data2, aes(x = Index, y = Diabetes_Ratio, label = State, size = Diabetes_Ratio, col = "red")) +
  geom_point() +
  #geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Diabetes Ratio vs Index",
       x = "Index",
       y = "Diabetes/Density") +
  scale_size_continuous(range = c(1, 10)) +
  theme_minimal() +
  geom_text(position = position_nudge(y = 0.05), size = 3, vjust = -2) +
  theme(plot.title = element_text(size = 15, face = "bold"),
        panel.grid.major = element_line(color = "gray", linetype = "solid"),
        panel.grid.minor = element_line(color = "gray", linetype = "solid"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))

ggplot(merged_data2, aes(x = Utilities, y = Diabetes_Ratio, label = State, size = Diabetes_Ratio, col = "red")) +
  geom_point() +
  #geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Diabetes Ratio vs Utilities",
       x = "Utilities",
       y = "Diabetes/Density") +
  scale_size_continuous(range = c(1, 10)) +
  theme_minimal() +
  geom_text(position = position_nudge(y = 0.05), size = 3, vjust = -2) +
  theme(plot.title = element_text(size = 15, face = "bold"),
        panel.grid.major = element_line(color = "gray", linetype = "solid"),
        panel.grid.minor = element_line(color = "gray", linetype = "solid"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))


ggplot(merged_data2, aes(x = Transportation, y = Diabetes_Ratio, label = State, size = Diabetes_Ratio, col = "red")) +
  geom_point() +
  #geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Diabetes Ratio vs Transportation",
       x = "Transportation",
       y = "Diabetes/Density") +
  scale_size_continuous(range = c(1, 10)) +
  theme_minimal() +
  geom_text(position = position_nudge(y = 0.05), size = 3, vjust = -2) +
  theme(plot.title = element_text(size = 15, face = "bold"),
        panel.grid.major = element_line(color = "gray", linetype = "solid"),
        panel.grid.minor = element_line(color = "gray", linetype = "solid"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))


ggplot(merged_data2, aes(x = Housing, y = Diabetes_Ratio, label = State, size = Diabetes_Ratio, col = "red")) +
  geom_point() +
  #geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Diabetes Ratio vs Housing",
       x = "Housing",
       y = "Diabetes/Density") +
  scale_size_continuous(range = c(1, 10)) +
  theme_minimal() +
  geom_text(position = position_nudge(y = 0.05), size = 3, vjust = -1.5) +
  theme(plot.title = element_text(size = 15, face = "bold"),
        panel.grid.major = element_line(color = "gray", linetype = "solid"),
        panel.grid.minor = element_line(color = "gray", linetype = "solid"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))

ggplot(merged_data2, aes(x = Grocery, y = Diabetes_Ratio, label = State, size = Diabetes_Ratio, col = "red")) +
  geom_point() +
  #geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Diabetes Ratio vs Grocery",
       x = "Grocery",
       y = "Diabetes/Density") +
  scale_size_continuous(range = c(1, 10)) +
  theme_minimal() +
  geom_text(position = position_nudge(y = 0.05), size = 3, vjust = -1.5) +
  theme(plot.title = element_text(size = 15, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))


ggplot(merged_data2, aes(x = Health, y = Diabetes_Ratio, label = State, size = Diabetes_Ratio, col = "red")) +
  geom_point() +
  #geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Diabetes Ratio vs Health",
       x = "Health",
       y = "Diabetes/Density") +
  scale_size_continuous(range = c(1, 10)) +
  theme_minimal() +
  geom_text(position = position_nudge(y = 0.05), size = 3, vjust = -1.5) +
  theme(plot.title = element_text(size = 15, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))



```



```{r }

diabetes<-diabetes_data


set.seed(123)

#split the dataset
split = sample.split(diabetes_data$Diabetes, 
                     SplitRatio = 0.80)
diabetes_train = subset(diabetes_data, 
                        split == TRUE)
diabetes_test = subset(diabetes_data, 
                       split == FALSE)

#other models,tree and naive since doesn't require scaling,will use train and test dataset
train<-diabetes_train
test<-diabetes_test


#Feature Scaling except target  for logistic regression and knn
diabetes_train[-22] = scale(diabetes_train[-22])
diabetes_test[-22] = scale(diabetes_test[-22])


#for upsampling
diabetes_train_up <- oversample(train, ratio = 0.8, method = "SMOTE",classAttr = "Diabetes")
diabetes_test_up<-test


#for other models create dataset,tree and naive since doesn't require scaling
train_up<-diabetes_train_up
test_up<-diabetes_test_up

#scale for upsampling
#Feature Scaling except target  for logistic regression and knn
diabetes_train_up[-22] = scale(diabetes_train_up[-22])
diabetes_test_up[-22] = scale(diabetes_test_up[-22])




#Model 1 logistic regression
logistic_model <- glm(Diabetes~.,data=diabetes_train,family="binomial")
summary(logistic_model)

#MI,Children,State Education is not significant. 

#lest see how the model predicts on the test data-probabilities
diabetes_test$prob<-predict(logistic_model, diabetes_test[,-22], type="response")


#lets create the predicted class
diabetes_test$predicted<-ifelse(diabetes_test$prob>=0.50,1,0)



Confusion_mat<- caret::confusionMatrix(as.factor(diabetes_test$predicted), as.factor(diabetes_test$Diabetes),positive='1')
Confusion_mat


draw_confusion_matrix <- function(cm,title) {
  
  layout(matrix(c(1,1,2)))
  par(mar=c(2,2,2,2))
  plot(c(100, 345), c(300, 450), type = "n", xlab="", ylab="", xaxt='n', yaxt='n')
  title(title, cex.main=2)
  
  # create the matrix 
  rect(150, 430, 240, 370, col='#3F97D0')
  text(195, 435, '0', cex=1.2)
  rect(250, 430, 340, 370, col='#F7AD50')
  text(295, 435, '1', cex=1.2)
  text(125, 370, 'Predicted', cex=1.3, srt=90, font=2)
  text(245, 450, 'Actual', cex=1.3, font=2)
  rect(150, 305, 240, 365, col='#F7AD50')
  rect(250, 305, 340, 365, col='#3F97D0')
  text(140, 400, '0', cex=1.2, srt=90)
  text(140, 335, '1', cex=1.2, srt=90)
  
  # add in the cm results 
  res <- as.numeric(cm$table)
  text(195, 400, res[1], cex=1.6, font=2, col='white')
  text(195, 335, res[2], cex=1.6, font=2, col='white')
  text(295, 400, res[3], cex=1.6, font=2, col='white')
  text(295, 335, res[4], cex=1.6, font=2, col='white')
  
  # add in the specifics 
  plot(c(100, 0), c(100, 0), type = "n", xlab="", ylab="", main = "DETAILS", xaxt='n', yaxt='n')
  text(10, 85, names(cm$byClass[1]), cex=1.2, font=2)
  text(10, 70, round(as.numeric(cm$byClass[1]), 3), cex=1.2)
  text(30, 85, names(cm$byClass[2]), cex=1.2, font=2)
  text(30, 70, round(as.numeric(cm$byClass[2]), 3), cex=1.2)
  text(50, 85, names(cm$byClass[5]), cex=1.2, font=2)
  text(50, 70, round(as.numeric(cm$byClass[5]), 3), cex=1.2)
  text(70, 85, names(cm$byClass[6]), cex=1.2, font=2)
  text(70, 70, round(as.numeric(cm$byClass[6]), 3), cex=1.2)
  text(90, 85, names(cm$byClass[7]), cex=1.2, font=2)
  text(90, 70, round(as.numeric(cm$byClass[7]), 3), cex=1.2)
  
  # add in the accuracy information 
  text(30, 35, names(cm$overall[1]), cex=1.5, font=2)
  text(30, 20, round(as.numeric(cm$overall[1]), 3), cex=1.4)
  text(70, 35, names(cm$overall[2]), cex=1.5, font=2)
  text(70, 20, round(as.numeric(cm$overall[2]), 3), cex=1.4)
} 

draw_confusion_matrix(Confusion_mat,'Logistic Confusion Matrix')



class(diabetes_test$prob)
class(diabetes_test$Diabetes)


predicts <- prediction(as.numeric(diabetes_test$prob),as.factor(diabetes_test$Diabetes))

roc <- performance(predicts,"tpr", "fpr")
plot(roc,main="ROC curve for GLM model")
auc_ROCR <- performance(predicts, measure = "auc")
auc_ROCR <- auc_ROCR@y.values[[1]]
print(paste("AUC value for logistic regression: ",auc_ROCR))


#run second model KNN
#extract 22 column of train dataset because we will use it as 'cl' argument in knn function.
target_category <- diabetes_train[,'Diabetes']


#extract 22th column if test dataset to measure the accuracy
test_category <- diabetes_test[,'Diabetes']


library(kknn)
library(class)
#lets run knn function
# getting k=317 since approach is to get square root of observations
pred <- knn(diabetes_train,diabetes_test[1:22],cl=target_category,k=317)
head(pred)



library(caret)
knn_Confusion_mat<- caret::confusionMatrix(as.factor(pred), as.factor(diabetes_test$Diabetes),positive='1')
knn_Confusion_mat

draw_confusion_matrix(knn_Confusion_mat,'KNN Confusion Matrix')


#show AUC for KNN
knn_predicts <- prediction(as.numeric(pred),as.numeric(diabetes_test$Diabetes))
roc <- performance(knn_predicts,"tpr", "fpr")
plot(roc,main="ROC curve for KNN model")
auc_ROCR_knn <- performance(knn_predicts, measure = "auc")
auc_ROCR_knn <- auc_ROCR_knn@y.values[[1]]
print(paste("AUC value for KNN Classification: ",auc_ROCR_knn))




#run Model3 Tree model

library(rpart)

library(rpart.plot) 

# build tree model
tree_model <- rpart(Diabetes ~ ., data = train, method = "class", control = rpart.control(cp = 0))

#https://www.statology.org/classification-and-regression-trees-in-r/
#view results
printcp(tree_model)

#identify best cp value to use
best <- tree_model$cptable[which.min(tree_model$cptable[,"xerror"]),"CP"]

#produce a pruned tree based on the best cp value
pruned_tree <- prune(tree_model, cp=best)

#plot the pruned tree
prp(pruned_tree,
    faclen=0, 
    extra=1,
    roundint=F,
    digits=5) 


# Make predictions on the test data
tree_pred <- predict(pruned_tree, test, type = "class")


tree_Confusion_mat<- caret::confusionMatrix(as.factor(tree_pred), as.factor(test$Diabetes),positive='1')
tree_Confusion_mat

draw_confusion_matrix(tree_Confusion_mat,'Tree Confusion Matrix')

#show AUC
treepredict <- prediction(as.numeric(tree_pred),as.numeric(test$Diabetes))
roc <- performance(treepredict,"tpr", "fpr")
plot(roc,main="ROC curve for GLM model")
tree_auc_ROCR <- performance(treepredict, measure = "auc")
tree_auc_ROCR <- tree_auc_ROCR@y.values[[1]]
print(paste("AUC value for Tree Classification: ",tree_auc_ROCR))



#Model 4 Naive Bayes

library(e1071) 
naive_model = naiveBayes(Diabetes ~., data = train)
summary(naive_model)

# predicting target 
nbc_pred <- test[1:21]
dim(nbc_pred)

naive_pred <- predict(naive_model, test)


naive_Confusion_mat<- caret::confusionMatrix(as.factor(naive_pred), as.factor(test$Diabetes),positive='1')
naive_Confusion_mat

draw_confusion_matrix(naive_Confusion_mat,'Naive Bayes Confusion Matrix')

#show AUC
naivepredict <- prediction(as.numeric(naive_pred),as.numeric(test$Diabetes))
roc <- performance(naivepredict,"tpr", "fpr")
plot(roc,main="ROC curve for GLM model")
naive_auc_ROCR <- performance(naivepredict, measure = "auc")
naive_auc_ROCR <- naive_auc_ROCR@y.values[[1]]
print(paste("AUC value for Naive Bayes Classification: ",naive_auc_ROCR))

#summary

draw_confusion_matrix(Confusion_mat,'Logistic Confusion Matrix')
print(paste("AUC value for logistic regression: ",round(auc_ROCR,3)))

draw_confusion_matrix(knn_Confusion_mat,'KNN Confusion Matrix')
print(paste("AUC value for KNN Classification: ",round(auc_ROCR_knn,3)))



draw_confusion_matrix(tree_Confusion_mat,'Tree Confusion Matrix')
print(paste("AUC value for Tree Classification: ",round(tree_auc_ROCR,3)))


draw_confusion_matrix(naive_Confusion_mat,'Naive Bayes Confusion Matrix')
print(paste("AUC value for Naive Bayes Classification: ",round(naive_auc_ROCR,3)))

#########################################################################################
#Now lets ran the models with upsampling data

#Model 1 logistic regression upsampling
logistic_model_up <- glm(Diabetes~.,data=diabetes_train_up,family="binomial")
summary(logistic_model_up)

#MI,Children,State Education is not significant. 

#lest see how the model predicts on the test data-probabilities
diabetes_test_up$prob<-predict(logistic_model_up, diabetes_test_up[,-22], type="response")


#lets create the predicted class
diabetes_test_up$predicted<-ifelse(diabetes_test_up$prob>=0.50,1,0)



Up_Confusion_mat<- caret::confusionMatrix(as.factor(diabetes_test_up$predicted), as.factor(diabetes_test_up$Diabetes),positive='1')
Up_Confusion_mat



draw_confusion_matrix(Up_Confusion_mat,'Upsampling Logistic Confusion Matrix')

predicts_up <- prediction(as.numeric(diabetes_test_up$prob),as.numeric(diabetes_test_up$Diabetes))
roc <- performance(predicts_up,"tpr", "fpr")
plot(roc,main="ROC curve for GLM model")
auc_ROCR_up <- performance(predicts_up, measure = "auc")
auc_ROCR_up <- auc_ROCR_up@y.values[[1]]
print(paste("AUC value for upsampling logistic regression: ",auc_ROCR_up))



#run model 2 KNN Upsampling
#extract 22 column of train dataset because we will use it as 'cl' argument in knn function.
target_category_up <- diabetes_train_up[,'Diabetes']


#extract 22th column if test dataset to measure the accuracy
test_category_up <- diabetes_test_up[,'Diabetes']


#lets run knn function
# getting k=317 since approach is to get square root of observations
pred_up <- knn(diabetes_train_up,diabetes_test_up[1:22],cl=target_category_up,k=317)


knn_Confusion_mat_up<- caret::confusionMatrix(as.factor(pred_up), as.factor(diabetes_test_up$Diabetes),positive='1')
knn_Confusion_mat_up


draw_confusion_matrix(knn_Confusion_mat_up,'Upsampling KNN  Confusion Matrix')


#show AUC for KNN
knn_predicts_up <- prediction(as.numeric(pred_up),as.numeric(diabetes_test_up$Diabetes))
roc <- performance(knn_predicts_up,"tpr", "fpr")
plot(roc,main="ROC curve for KNN model")
auc_ROCR_knn_up <- performance(knn_predicts_up, measure = "auc")
auc_ROCR_knn_up <- auc_ROCR_knn_up@y.values[[1]]
print(paste("AUC value for KNN Classification: ",auc_ROCR_knn_up))



#build 3.model
# build tree model
tree_model_up <- rpart(Diabetes ~ ., data = train_up, method = "class", control = rpart.control(cp = 0))

#show important variables
varImp(tree_model_up,surrogates = FALSE, competes = FALSE)

# Make predictions on the test data
tree_pred_up <- predict(tree_model_up, test_up, type = "class")


tree_Confusion_mat_up<- caret::confusionMatrix(as.factor(tree_pred_up), as.factor(test_up$Diabetes),positive='1')
tree_Confusion_mat_up

draw_confusion_matrix(tree_Confusion_mat_up,'Upsampling Tree Confusion Matrix')

#show AUC
treepredict_up <- prediction(as.numeric(tree_pred_up),as.numeric(test_up$Diabetes))
roc <- performance(treepredict_up,"tpr", "fpr")
plot(roc,main="ROC curve for GLM model")
tree_auc_ROCR_up <- performance(treepredict_up, measure = "auc")
tree_auc_ROCR_up <- tree_auc_ROCR_up@y.values[[1]]
print(paste("AUC value for Upsampling Tree Classification: ",tree_auc_ROCR_up))


#Model 4 Naive Bayes
library(e1071) 
naive_model_up = naiveBayes(Diabetes ~., data = train_up)
summary(naive_model_up)

# predicting target 
nbc_pred_up <- test_up[1:21]
dim(nbc_pred_up)

naive_pred_up <- predict(naive_model_up, test_up)


naive_Confusion_mat_up<- caret::confusionMatrix(as.factor(naive_pred_up), as.factor(test_up$Diabetes),positive='1')
naive_Confusion_mat_up

draw_confusion_matrix(naive_Confusion_mat_up,'Upsampling Naive Bayes Confusion Matrix')

#show AUC
naivepredict_up <- prediction(as.numeric(naive_pred_up),as.numeric(test_up$Diabetes))
roc <- performance(naivepredict_up,"tpr", "fpr")
plot(roc,main="ROC curve for GLM model")
naive_auc_ROCR_up <- performance(naivepredict_up, measure = "auc")
naive_auc_ROCR_up <- naive_auc_ROCR_up@y.values[[1]]
print(paste("AUC value for Upsampling Naive Bayes Classification: ",naive_auc_ROCR_up))




#summary for upsampling
draw_confusion_matrix(Up_Confusion_mat,'Upsampling Logistic Confusion Matrix')
print(paste("AUC value for Upsamling logistic regression: ",round(auc_ROCR_up,3)))


draw_confusion_matrix(knn_Confusion_mat_up,'Upsampling KNN Confusion Matrix')
print(paste("AUC value for Upsampling KNN Classification: ",round(auc_ROCR_knn_up,3)))


draw_confusion_matrix(tree_Confusion_mat_up,'Upsampling Tree Confusion Matrix')
print(paste("AUC value for Upsampling Tree Classification: ",round(tree_auc_ROCR_up,3)))


draw_confusion_matrix(naive_Confusion_mat_up,' Upsampling Naive Bayes Confusion Matrix')
print(paste("AUC value for Upsampling Naive Bayes Classification: ",round(naive_auc_ROCR_up,3)))

```
